---
title: "Midterm Project (Classification Problem)"
author: "Mithoon Chakraborty"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Loading Dataset
```{r}
adult <- read.table(
  "adult.data",
  sep = ",",
  header = FALSE,
  strip.white = TRUE,
  stringsAsFactors = FALSE,
  colClasses = "character"
)

colnames(adult) <- c(
  "age", "workclass", "fnlwgt", "education", "education_num",
  "marital_status", "occupation", "relationship", "race", "sex",
  "capital_gain", "capital_loss", "hours_per_week", "native_country", "income"
)

dim(adult)
head(adult)
```

# 2. Data Preprocessing & Exploration
## 2.1 Exploratory Data Analysis
```{r}
# view raw data
summary(adult)
# checking structure
str(adult)
# variable distribution
table(adult$income)
prop.table(table(adult$income))

# Income Class Distribution
barplot(table(adult$income),
        main = "Income Class Distribution",
        xlab = "Income Category",
        ylab = "Frequency",
        col = c("lightblue", "lightgreen"))

# Frequency tables (categorical variables)
table(adult$workclass)
table(adult$education)
table(adult$marital_status)
table(adult$occupation)
table(adult$relationship)
table(adult$race)
table(adult$sex)
table(adult$native_country)
table(adult$income)
# Histograms (numeric variables)
hist(as.numeric(adult$age), main = "Age Distribution", xlab = "Age")

hist(as.numeric(adult$fnlwgt), main = "Fnlwgt Distribution", xlab = "Fnlwgt")

hist(as.numeric(adult$education_num), main = "Education Num Distribution", xlab = "Education Num")

hist(as.numeric(adult$capital_gain), main = "Capital Gain Distribution", xlab = "Capital Gain")

hist(as.numeric(adult$capital_loss), main = "Capital Loss Distribution", xlab = "Capital Loss")

hist(as.numeric(adult$hours_per_week), main = "Hours per Week Distribution", xlab = "Hours")

# Numeric Summary 
numeric_data <- adult[, c("age","fnlwgt","education_num","capital_gain","capital_loss","hours_per_week")]

data.frame(
  Variable = names(numeric_data),
  Mean = sapply(numeric_data, function(x) mean(as.numeric(x))),
  Median = sapply(numeric_data, function(x) median(as.numeric(x))),
  Min = sapply(numeric_data, function(x) min(as.numeric(x))),
  Max = sapply(numeric_data, function(x) max(as.numeric(x)))
)

# corelation matrix
numeric_data2 <- data.frame(lapply(numeric_data, as.numeric))
round(cor(numeric_data2), 2)

# Bar plots (categorical variables)
barplot(table(adult$income), main = "Income Distribution")

barplot(table(adult$sex), main = "Sex Distribution")

barplot(table(adult$workclass), main = "Workclass Distribution")

barplot(table(adult$education), main = "Education Distribution")

# Boxplots for Outlier Detection
boxplot(as.numeric(adult$age), main = "Boxplot of Age")

boxplot(as.numeric(adult$fnlwgt), main = "Boxplot of Fnlwgt")

boxplot(as.numeric(adult$capital_gain), main = "Boxplot of Capital Gain")

boxplot(as.numeric(adult$capital_loss), main = "Boxplot of Capital Loss")

boxplot(as.numeric(adult$hours_per_week), main = "Boxplot of Hours per Week")


# Missing value distribution
colSums(adult == "?")

```
## 2.2 Data Preprocessing
### 2.2.1 Handling Missing Value
```{r}
# Count missing values
colSums(adult == "?")
# Store missing counts
missing_counts <- colSums(adult == "?")
# Missing Percentage
missing_percent <- missing_counts / nrow(adult) * 100
missing_percent
# Convert missing value to NA
adult[adult == "?"] <- NA
# Verify missing value
colSums(is.na(adult))
# Cleaning missing value
adult_clean <- na.omit(adult)
# Compare dataset size before and after cleaning
dim(adult)
dim(adult_clean)
# Number of rows removed
rows_removed <- nrow(adult) - nrow(adult_clean)
rows_removed
```
### 2.2.2 Handling Outliers
```{r}
# Outlier Handling Decision
# Outliers identified during EDA
# No removal performed
# Boxplot for Outlier Detection
boxplot(as.numeric(adult_clean$capital_gain),
        main = "Boxplot of Capital Gain",
        col = "lightgreen")
boxplot(as.numeric(adult_clean$capital_loss),
        main = "Boxplot of Capital Loss",
        col = "lightblue")
```
### 2.2.3 Data Transformation
```{r}
# Convert numeric variables
adult_clean$age <- as.numeric(adult_clean$age)
adult_clean$fnlwgt <- as.numeric(adult_clean$fnlwgt)
adult_clean$education_num <- as.numeric(adult_clean$education_num)
adult_clean$capital_gain <- as.numeric(adult_clean$capital_gain)
adult_clean$capital_loss <- as.numeric(adult_clean$capital_loss)
adult_clean$hours_per_week <- as.numeric(adult_clean$hours_per_week)

# Convert categorical variables
factor_cols <- c(
  "workclass", "education", "marital_status", "occupation",
  "relationship", "race", "sex", "native_country", "income"
)
adult_clean[factor_cols] <- lapply(adult_clean[factor_cols], as.factor)

# Verify Data Types
str(adult_clean)

# Handle Skewness using Log Transformation
adult_clean$capital_gain_log <- log1p(adult_clean$capital_gain)
adult_clean$capital_loss_log <- log1p(adult_clean$capital_loss)
adult_clean$fnlwgt_log <- log1p(adult_clean$fnlwgt)

```
## 2.3 Relationship Visualization
```{r}
# Age vs Income
boxplot(as.numeric(age) ~ income, data = adult_clean,
        main = "Age vs Income", xlab = "Income", ylab = "Age")

# Hours per week vs Income
boxplot(as.numeric(hours_per_week) ~ income, data = adult_clean,
        main = "Hours per Week vs Income", xlab = "Income", ylab = "Hours")

# Education vs Income
prop.table(table(adult_clean$education, adult_clean$income), 1)

# Sex vs Income
prop.table(table(adult_clean$sex, adult_clean$income), 1)

# Occupation vs Income
table(adult_clean$occupation, adult_clean$income)
```
# 3. Model Development: 
## 3.1 Preparing The modeling dataset
```{r}
# Final Modeling Dataset Check
str(adult_clean)
summary(adult_clean)

# Target Variable Check
table(adult_clean$income)
prop.table(table(adult_clean$income))
```

## 3.2 Split the dataset
```{r}
# Training and Validation Split
set.seed(123)

train_index <- sample(1:nrow(adult_clean), 0.7 * nrow(adult_clean))

train_data <- adult_clean[train_index, ]
valid_data <- adult_clean[-train_index, ]

dim(train_data)
dim(valid_data)

```

## 3.3 Choose predictor variables
```{r}
income ~ age + workclass + education + education_num + marital_status +
occupation + relationship + race + sex + capital_gain + capital_loss +
hours_per_week + native_country
```

## 3.4 Model 1 — Logistic Regression
```{r}
# Logistic Regression Model using resampling
library(caret)
ctrl <- trainControl(method = "boot", number = 25)

# Logistic Regression Model
log_model <- train(
  income ~ age + workclass + education + education_num + marital_status +
    occupation + relationship + race + sex + capital_gain + capital_loss +
    hours_per_week + native_country,
  data = train_data,
  method = "glm",
  family = "binomial",
  trControl = ctrl
)

log_model

# Coefficient Table
coef_table <- round(summary(log_model$finalModel)$coefficients, 4)
coef_table

# Show only significant variables
coef_table[coef_table[,4] < 0.05, ]

# Pseudo R²
pseudo_r2 <- 1 - (log_model$finalModel$deviance / log_model$finalModel$null.deviance)
round(pseudo_r2, 4)

# AIC
log_model$finalModel$aic

# Predictions
log_pred <- predict(log_model, newdata = valid_data)

# Confusion Matrix
cm_log <- confusionMatrix(log_pred, valid_data$income)
cm_log

# Accuracy
log_accuracy <- round(cm_log$overall["Accuracy"], 4)
log_accuracy

# Precision
log_precision <- round(cm_log$byClass["Pos Pred Value"], 4)
log_precision

# Recall
log_recall <- round(cm_log$byClass["Sensitivity"], 4)
log_recall

# F1 Score
log_f1 <- round(2 * (log_precision * log_recall) / (log_precision + log_recall), 4)
log_f1
```

## 3.5 Model 2 — Linear Discriminant Analysis
```{r}
# LDA Model using caret
library(caret)

lda_model <- train(
  income ~ age + workclass + education + marital_status + occupation +
    relationship + race + sex + capital_gain + capital_loss +
    hours_per_week + native_country,
  data = train_data,
  method = "lda"
)

lda_model

# LDA model Prediction
lda_pred <- predict(lda_model, newdata = valid_data)

# LDA model confusion matrix
cm_lda <- confusionMatrix(lda_pred, valid_data$income)
cm_lda

# Accuracy
lda_accuracy <- round(cm_lda$overall["Accuracy"], 4)
lda_accuracy

# Precision
lda_precision <- round(cm_lda$byClass["Pos Pred Value"], 4)
lda_precision

# Recall
lda_recall <- round(cm_lda$byClass["Sensitivity"], 4)
lda_recall

# F1 Score
lda_f1 <- round(2 * (lda_precision * lda_recall) / (lda_precision + lda_recall), 4)
lda_f1
```
## 3.6 Model 3 — Decission Tree
```{r}
library(caret)

# Decision Tree Model using caret
tree_model <- train(
  income ~ age + workclass + education + marital_status +
    occupation + relationship + race + sex + capital_gain + capital_loss +
    hours_per_week + native_country,
  data = train_data,
  method = "rpart"
)

tree_model

# Decision Tree Prediction
tree_pred <- predict(tree_model, newdata = valid_data)

# Decision Tree confusion matrix
cm_tree <- confusionMatrix(tree_pred, valid_data$income)
cm_tree

# Accuracy
tree_accuracy <- round(cm_tree$overall["Accuracy"], 4)
tree_accuracy

# Precision
tree_precision <- round(cm_tree$byClass["Pos Pred Value"], 4)
tree_precision

# Recall
tree_recall <- round(cm_tree$byClass["Sensitivity"], 4)
tree_recall

# F1 Score
tree_f1 <- round(2 * (tree_precision * tree_recall) / (tree_precision + tree_recall), 4)
tree_f1
```
## 3.7 Model Comparison
```{r}
model_comparison <- data.frame(
  Model = c("Logistic Regression", "LDA", "Decision Tree"),
  Accuracy = c(log_accuracy, lda_accuracy, tree_accuracy),
  Precision = c(log_precision, lda_precision, tree_precision),
  Recall = c(log_recall, lda_recall, tree_recall),
  F1_Score = c(log_f1, lda_f1, tree_f1)
)

knitr::kable(model_comparison, caption = "Model Comparison on Validation Data")
```
# 4. Performance Metrics and Testing Selected Logistic Model on Test Data set
## 4.1 Model Performance Metrics on Validation Data
```{r}
# Final Model Performance (Validation Set)
library(caret)

# Predictions
val_pred <- predict(log_model, newdata = valid_data)

# Confusion Matrix (Full metrics)
conf_matrix <- confusionMatrix(val_pred, valid_data$income)
conf_matrix

# Accuracy
val_accuracy <- round(conf_matrix$overall["Accuracy"], 4)
val_accuracy

# Precision
val_precision <- round(conf_matrix$byClass["Pos Pred Value"], 4)
val_precision

# Recall (Sensitivity)
val_recall <- round(conf_matrix$byClass["Sensitivity"], 4)
val_recall

# F1 Score
val_f1 <- round(2 * (val_precision * val_recall) / (val_precision + val_recall), 4)
val_f1

# Combine all metrics in one table
performance_metrics <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
  Value = c(val_accuracy, val_precision, val_recall, val_f1)
)

performance_metrics
```
## 4.2 Model Performance Metrics on Test Data
### 4.2.1 LoadTest Dataset and Preprocessing
```{r}

# Load Test Dataset

adult_test <- read.table(
  "adult.test",
  sep = ",",
  header = FALSE,
  strip.white = TRUE,
  stringsAsFactors = FALSE,
  colClasses = "character",
  skip = 1
)

colnames(adult_test) <- colnames(adult)


# Clean Test Dataset
# Remove "." from income labels
adult_test$income <- gsub("\\.", "", adult_test$income)

# Convert "?" to NA
adult_test[adult_test == "?"] <- NA

# Remove missing values
adult_test_clean <- na.omit(adult_test)

# Match Data Types with Training Data


# Numeric variables
adult_test_clean$age <- as.numeric(adult_test_clean$age)
adult_test_clean$fnlwgt <- as.numeric(adult_test_clean$fnlwgt)
adult_test_clean$education_num <- as.numeric(adult_test_clean$education_num)
adult_test_clean$capital_gain <- as.numeric(adult_test_clean$capital_gain)
adult_test_clean$capital_loss <- as.numeric(adult_test_clean$capital_loss)
adult_test_clean$hours_per_week <- as.numeric(adult_test_clean$hours_per_week)

# Categorical variables
factor_cols <- c(
  "workclass", "education", "marital_status", "occupation",
  "relationship", "race", "sex", "native_country", "income"
)

adult_test_clean[factor_cols] <- lapply(adult_test_clean[factor_cols], factor)

# Make factor levels match training data
adult_test_clean$workclass <- factor(adult_test_clean$workclass, levels = levels(train_data$workclass))
adult_test_clean$education <- factor(adult_test_clean$education, levels = levels(train_data$education))
adult_test_clean$marital_status <- factor(adult_test_clean$marital_status, levels = levels(train_data$marital_status))
adult_test_clean$occupation <- factor(adult_test_clean$occupation, levels = levels(train_data$occupation))
adult_test_clean$relationship <- factor(adult_test_clean$relationship, levels = levels(train_data$relationship))
adult_test_clean$race <- factor(adult_test_clean$race, levels = levels(train_data$race))
adult_test_clean$sex <- factor(adult_test_clean$sex, levels = levels(train_data$sex))
adult_test_clean$native_country <- factor(adult_test_clean$native_country, levels = levels(train_data$native_country))
adult_test_clean$income <- factor(adult_test_clean$income, levels = levels(train_data$income))

# Check structure
str(adult_test_clean)

# Check
dim(adult_test_clean)
head(adult_test_clean)
```

### 4.2.2 Predict using FINAL MODEL (Logistic)
```{r}
# Test Prediction (Final Model)
test_pred <- predict(log_model, newdata = adult_test_clean)
```
### 4.2.3 Model Confusion Matrix on Test Data
```{r}
# Test Confusion Matrix
conf_matrix_test <- confusionMatrix(test_pred, adult_test_clean$income)
conf_matrix_test
```
### 4.2.4 Model Accuracy on Test Data
```{r}
# Test Accuracy
test_accuracy <- round(mean(test_pred == adult_test_clean$income), 4)
test_accuracy

# Extract full confusion matrix
cm_test <- confusionMatrix(test_pred, adult_test_clean$income)

# Accuracy
test_accuracy <- round(cm_test$overall["Accuracy"], 4)

# Precision
test_precision <- round(cm_test$byClass["Pos Pred Value"], 4)

# Recall (Sensitivity)
test_recall <- round(cm_test$byClass["Sensitivity"], 4)

# F1 Score
test_f1 <- round(2 * (test_precision * test_recall) / (test_precision + test_recall), 4)

# Create table
test_performance <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
  Value = c(test_accuracy, test_precision, test_recall, test_f1)
)

test_performance
```



